---
title: "Homework: Chapter 8 and Model Comparison"
format: html
author: "Kaylynn Hiller"
date: "2026-03-23"
---

## Setup

Use the same `states` data from the chapter, with the `inc1000` variable.

```{r}
#| eval: false
pacman::p_load(dplyr, tidyr, broom, stringr, modelsummary, tinyplot, ggplot2)

state.x77_with_names <- as_tibble(state.x77, rownames = "state")

states <- bind_cols(state.x77_with_names,
                    region = state.division) |>
  janitor::clean_names() |>
  mutate(south = as.integer(str_detect(as.character(region), "South")),
         inc1000 = income / 1000)
```

## The Additive Model

Estimate this additive model and display the results with `msummary()`:

```{r}
#| eval: false
m1 <- lm(life_exp ~ hs_grad + inc1000, data = states)
msummary(m1)
```

Write a sentence interpreting each of the two slope coefficients. Use appropriately hedged language (e.g., "adjusting for...").

-   For every one one percent increase in the high school graduation rate in a US state, adjusting for per capita income in thousands of USD, we expect average life expectancy in a US state to increase by 0.1 years.
-   For every one one unit increase in per capita income in thousands of USD in a US state, adjusting for the high school graduation rate, we expect average life expectancy in a US state to decrease by 0.073 years.

Now compute and examine the 95% confidence interval for `inc1000`. What does it tell you about the precision of that estimate? What might this imply about whether income is a useful predictor of life expectancy after accounting for education?

```{r}
tidy(m1, conf.int = TRUE, conf.level = .95)

b2 <- coef(m1)["inc1000"]
SE <- summary(m1)$coefficients["inc1000", "Std. Error"]
tcrit <- qt(0.975, df = 47)
CI_low <- b2 - (tcrit*SE)
CI_high <- b2 + (tcrit*SE)
```

-   (-0.737, 0.591)

-   The precision of this estimate is lower compared to the high school graduation rate as the interval is much larger. Additionally, the confidence interval contains zero and the associated p-value is greater than 0.05 which suggests that income in not a useful predictor of life expectancy after accounting for the high school graduation rate.

## Visualizing the Additive Model

The companion shows two ways to plot `m1`: with `hs_grad` on the x-axis at representative values of income, and then with axes swapped. Make one of these plots (your choice).

```{r}
# three values of income
myincs <- quantile(states$inc1000,
                   c(0, .5, 1))

# this will be the x-axis
myhs <- seq(min(states$hs_grad),
            max(states$hs_grad),
            length = 100)

# combine the values
xgrid <- expand_grid(inc1000 = myincs,
                     hs_grad = myhs)

# get predictions and confidence bounds
yhat <- predict(m1,
                newdata = xgrid,
                interval = "confidence",
                level = .95)

# stick everything together and label the lines
preds <- bind_cols(yhat, xgrid) |>
  mutate(income = case_match(inc1000,
                             min(states$inc1000) ~ "min",
                             max(states$inc1000) ~ "max",
                             .default = "median"))

ggplot(preds, aes(x = hs_grad, y = fit, color = income, fill = income)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  labs(
    title = "Predicted life expectancy",
    subtitle = "Additive model, 95% CIs",
    y = "Life expectancy",
    x = "% of HS grad",
    color = "income",
    fill = "income")
```

In an additive model, the prediction lines for different values of the other variable are always parallel. Explain in 2--3 sentences what the parallelism means substantively. What assumption is the additive model making about how `hs_grad` and `inc1000` jointly affect life expectancy?

-   This parallelism means that this model assumes the the effect of high school graduation rate and per capita income are additive effects. The marginal effect of each of these outcomes on life expectancy does not depend on the other variable. This means that the amount that life expectancy increases by increasing graduation rate does not depend on the per capita income and vice versa.

## The Interaction Model

Now estimate:

```{r}
#| eval: false
m2 <- lm(life_exp ~ hs_grad * inc1000, data = states)

msummary(list(m1, m2))
```

Display `m1` and `m2` side by side using `msummary()`.

The interaction coefficient is negative. What does that mean? Describe the interaction from both directions: what does it say about the slope of `hs_grad` at different levels of income, and what does it say about the slope of `inc1000` at different levels of education?

-   A negative interaction coefficient means that as high school graduation rate increases, the effects of income on life expectancy decreases. This also means that as income increases, the effect of high school graduation rate on life expectancy also decreases.

## Visualizing the Interaction

Make the interaction plot from the companion --- predicted `life_exp` against `hs_grad` at three representative values of `inc1000` (minimum, median, maximum). Use either tinyplot or ggplot2.

```{r}
# three values of income
myincs <- quantile(states$inc1000,
                   c(0, .5, 1))

# this will be the x-axis
myhs <- seq(min(states$hs_grad),
            max(states$hs_grad),
            length = 100)

# combine the values
xgrid <- expand_grid(inc1000 = myincs,
                     hs_grad = myhs)

# get predictions and confidence bounds
yhat <- predict(m2,
                newdata = xgrid,
                interval = "confidence",
                level = .95)

# stick everything together and label the lines
preds <- bind_cols(yhat, xgrid) |>
  mutate(income = case_match(inc1000,
                             min(states$inc1000) ~ "min",
                             max(states$inc1000) ~ "max",
                             .default = "median"))

ggplot(preds, aes(x = hs_grad, y = fit, color = income, fill = income)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.2, color = NA) +
  geom_line(linewidth = 1) +
  labs(
    title = "Predicted life expectancy",
    subtitle = "Interactive model, 95% CIs",
    y = "Life expectancy",
    x = "% of HS grad",
    color = "income",
    fill = "income")
```

What's the key visual difference between this plot and the additive model plot? In the footnote to this plot, the companion hints that Alaska and Utah might be driving the interaction. What would you want to investigate if this were a real analysis?

-   The main difference is that rather than the slopes being the same, they are different. This means that at different income levels, the effect of graduation rate on life expectancy differs. We would want to investigate and examine Alaska and Utah to see if they are outliers.

## Comparing the Models

Use `anova(m1, m2)` to run the incremental F-test.

```{r}
anova(m1, m2)
```

What is the null hypothesis of this test? What do you conclude from the result?

-   The null hypothesis is that beta 3, the interaction coefficient, is zero. We would reject the null hypothesis based on the results of this anova.

Now run the likelihood ratio test with `lmtest::lrtest(m1, m2)`. The p-value from the LRT is smaller than the one from the F-test. The companion explains why --- state that explanation in your own words.

```{r}
lmtest::lrtest(m1, m2)
```

-   The likelihood test and ANOVA use different distributions to examine whether a statistic is surprising based on what would happen based on random chance. The likelihood test uses a chi squared distribution and ANOVA uses an F distribution. The tails of the chi squared distribution are thinner so the associated probability is smaller. This is because the F distribution is based on the degrees of freedom and the chi squared uses infinite degrees of freedom.

## Likelihood and Deviance

*These questions draw on the "Likelihood comparison" bonus section.*

Using `logLik()`, extract the log-likelihoods of `m1` and `m2`. Compute their deviances ($D = -2\ell$) by hand.

```{r}
dev_1 <- -2 * as.numeric(logLik(m1))
dev_2 <- -2 * as.numeric(logLik(m2))
dev_1
dev_2
```

The companion notes that `deviance()` on an `lm` object does *not* give you $-2\ell$ --- it gives you the RSS. Confirm this by comparing `deviance(m1)` to your hand-calculated $D$ for `m1`. Why are they different? What does each one measure?

```{r}
rss_1 <- deviance(m1)
rss_2 <- deviance(m2)
```

-   The deviance function on an lm model returns the residual sum of squares. Using our previous method, we found the deviance.

Compute the proportional reduction in deviance ($R^2$) for `m1` using both the RSS formula and the $\Delta D$ formula from the companion. Confirm they give the same answer.

```{r}
1 - exp( -(dev_1 - dev_2 ) / nrow(states))
(rss_1 - rss_2) / rss_1 
```

## AIC and BIC

*These questions draw on the "Information criteria" bonus section.*

Estimate a null model (`m0`), the additive model (`m1`), and the interaction model (`m2`). Then compute AIC and BIC for all three.

```{r}
#| eval: false
m0 <- lm(life_exp ~ 1, data = states)

AIC(m0, m1, m2)
BIC(m0, m1, m2)
```

Which model is preferred by AIC? Which by BIC? Are they the same?

-   Both prefer m2.

The companion explains that the AIC penalty of $2k$ corresponds to the expected increase in out-of-sample deviance per extra parameter. In your own words, what problem is the penalty term meant to solve? Why can't we just pick the model with the lowest deviance (i.e., the best in-sample fit)?

-   We want to decrease deviance without over fitting the model and adding junk parameters. The goal of statistics is to build models to be able to interpret our data. The model with the lowest deviance would include all possible parameters, but would not meaningfully tell us anything.

The BIC penalty is $k \log n$ instead of $2k$. For $n = 50$, compute $\log(50)$ and compare it to 2. What does this tell you about how the two criteria will differ in their preferences when $n$ is small vs. large?

-   Log(50) is around 3.9. This is larger than 2. This suggests that for smaller sample sizes, BIC penalizes adding additional parameters more than in larger sample sizes.
