---
title: "Week 2 Exercises"
author: "Kaylynn Hiller"
date: "2026-01-16"
mainfont: "Georgia"
format: 
  html: 
    theme: cosmo
    toc: true
    embed-resources: true
  pdf:
    colorlinks: true
    includes:
      in-header:
    keep-tex: true
header-includes:
  \usepackage{tabularray}
  \usepackage{siunitx}
execute: 
  echo: false
---

## Week 2

```{r}

library(nycflights13)
library(tidyverse)
```

**Exercise 3.2.5**

1.  In a single pipeline for each condition, find all flights that meet the condition:

-   Had an arrival delay of two or more hours

```{r}
flights |> 
  filter(arr_delay >= 120)
```

-   Flew to Houston (`IAH` or `HOU`)

```{r}
flights |> 
  filter(dest == "IAH" | dest == "HOU")
```

-   Were operated by United, American, or Delta

```{r}
flights |> 
  filter(carrier %in% c("UA", "AA", "DL"))
```

-   Departed in summer (July, August, and September)

```{r}
flights |> 
  filter(month %in% 7:9)
```

-   Arrived more than two hours late but didn’t leave late

```{r}
flights |> 
  filter(arr_delay > 120 & dep_delay <= 0)
```

-   Were delayed by at least an hour, but made up over 30 minutes in flight

```{r}
flights |> 
  filter(dep_delay >= 60 & dep_delay - arr_delay > 30)
```

2.  Sort flights to find the flights with the longest departure delays. Find the flights that left earliest in the morning.

```{r}
flights |> 
  arrange(desc(dep_delay))

flights |> 
  arrange(dep_time)
```

3.  Sort flights to find the fastest flights. (Hint: Try including a math calculation inside of your function.)

```{r}
to_kmh <- function(distance, minutes) {
  (distance * 1.60934) / (minutes / 60)
}

flights |> 
  mutate(speed = to_kmh(distance, air_time)) |>
  relocate(speed) |> 
  arrange(desc(speed))
```

4.  Was there a flight on every day of 2013? Yes

```{r}
flights |> 
  distinct(month, day) |> 
  summarize(num_days = n())
```

5.  Which flights traveled the farthest distance? Which traveled the least distance?

```{r}
flights |> 
  arrange(desc(distance))

flights |> 
  arrange(distance)
```

6.  Does it matter what order you used filter() and arrange() if you’re using both? Why/why not? Think about the results and how much work the functions would have to do.

It does not matter. Filter pulls out rows that meet a condition but does not change the ordering. Arrange changes the ordering but leaves all rows in. Since they perform separate functions, the order does not matter.

**Exercise 3.3.5**

1.  Compare dep_time, sched_dep_time, and dep_delay. How would you expect those three numbers to be related?

    I would expect the scheduled departure time (sched_dep_time) plus the delay (dep_delay) to equal the actual departure time (dep_time).

```{r}
flights |>
  select(dep_time, sched_dep_time, dep_delay)
```

2.  Brainstorm as many ways as possible to select dep_time, dep_delay, arr_time, and arr_delay from flights.
```{r}
flights

flights |>
  select(dep_time, dep_delay, arr_time, arr_delay)

flights |>
  select(dep_time:arr_delay)

flights |>
  select(!year:day & !sched_dep_time & !sched_arr_time &!carrier:time_hour)

flights |> 
  select(starts_with("dep_"), starts_with("arr_"))
```

3.  What happens if you specify the name of the same variable multiple times in a select() call?
```{r}
flights|>
  select(year, year)
```
You just get the variable once.

4.  What does the any_of() function do? Why might it be helpful in conjunction with this vector?
```{r}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")
flights|>
  select(any_of(variables))
```
The any_of() function in conjuction with the select function selects any variables listed. If you set specific variables to select outside the pipeline, the code is cleaner and easier to change.

5.  Does the result of running the following code surprise you? How do the select helpers deal with upper and lower case by default? How can you change that default?
```{r}
flights |> select(contains("TIME"))
```
The result did not surprise me. By default, select helpers treat upper and lower case the same. You can change this by specifying with ignore.case = FALSE.

6.  Rename air_time to air_time_min to indicate units of measurement and move it to the beginning of the data frame.
```{r}
flights |>
  relocate(air_time_min = air_time)
```

7.  Why doesn’t the following work, and what does the error mean?
```{r}
flights |> 
  select(tailnum) |> 
  arrange(arr_delay)
#> Error in `arrange()`:
#> ℹ In argument: `..1 = arr_delay`.
#> Caused by error:
#> ! object 'arr_delay' not found
```

We did not select that variable in the select function so it is not available to arrange.

**Exercise 3.5.7**

```{r}

```
